The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-01-02 pkg/partykit/inst/guideliketest/draft/draft.Rnw
revised sections 2 and 3 and fixed typos in references
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2399 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ffc676ddf3697959995a7bf4905fea229d54cf55

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
source("../guidelike.R")
source("../tests.R")
#load("../sim/simres20180422_1step.rda")
load("../sim/simres20181122_stump.rda")
sim_stump <- simres$res
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
load("../sim/simres20181122_tree.rda")
sim_tree <- simres$res
rm(simres)

## HCL palette
pal <- hcl(c(140, 80, 260, 50, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")

@@

\Abstract{
A core step of every algorithm for learning regression trees is the decision if, 
how, and where to split the underlying data - or, in other words, the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employ greedy search strategies 
that directly compare all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selection of covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using p~-~values that are appropriately adjusted for 
the possible split points. In a second step a split point optimizing some objective 
function is selected in the chosen split variable. However, different unbiased tree 
algorithms employ different inference frameworks for computing these p~-~values and 
their relative advantages or disadvantages are not well understood, yet.

Therefore, three different approaches are considered here and embedded into a common 
modeling framework with special emphasis to linear model trees: classical categorical 
association tests (GUIDE), conditional inference (CTree), parameter instability 
tests (MOB). It is assessed how different building blocks affect the power of the 
tree algorithms to select the appropriate covariates for splitting: residuals vs. full 
model scores, binarization of residuals/scores at zero, binning of covariates, 
conditional vs. unconditional approximations of the null distribution.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with high/extreme variations 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on (a set of) covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is minimized/maximized, they differ in their 
specific/particular approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all follow the idea of 
choosing the covariate which shows the highest association to/influence on the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory. 
In this paper the focus will be put on this first step, i.e. the task of selecting 
the right/best split variable. In particular, the approach of the GUIDE algorithm is
compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
unifying framework for unbiased model based tree algorithms is presented such that each 
of the three algorithms can be obtained by a specific combination of the available building 
blocks. In that way, the effects of each of these characteristic features of the GUIDE,
CTree and MOB algorithm can be investigated separately by varying the switches in
the tool box provided by this general framework.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated testing strategies more in detail in 
Section~\ref{sec:strategies} the unifying framework for model based unbiased
recursive partitioning is presented in Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

\section{Unbiased Recursive Partitioning}
\label{sec:urp}
The basic idea of building a regression tree model is to partition the data into smaller and more 
homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
have been developed, all following the same general structure:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
%This can for example be the sum of deviations from a typical or average value of the 
%data set or it can be based on a fitted model.
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data the discrepancy measure is minimal within the resulting subgroups.
\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criterion
is reached or the subgroup consists of only one or all equal observations. (In the second
case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
\end{enumerate}

@@

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical (independence) tests
following the same general strategy: 
Based on the discrepancy measure an $n \times k$ - matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
necessary to calculate the corresponding p~-~values in order to allow for an unbiased
variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
p~-~value is selected as split variable as it has shown to be the variable with the strongest
influence on the goodness of fit of the model.
% Hinweis: pruning-Strategien
%Once the split variable is chosen the corresponding split point is then selected such that it 
%minimizes the discrepancy measure.

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Section~\ref{sec:strategies}.

@@

\subsection{Testing strategies}\label{sec:strategies}

\subsubsection{CTree}\label{sec:ctree}
The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
a conditional inference framework by applying permutation tests. In its basic form linear test 
statistics are applied to measure the dependency between a covariate and the response variable. 
Optionally, maximally selected statistics can be employed as well. By allowing for transformations 
of the covariates and the application of so~-~called influence functions on the response variable 
within the test statistic a model-based approach is also possible. For example, the influence
function can calculate a model-based measure of goodness of fit as a discrepancy measure 
for each response value (in a permutation symmetric way) such that the test statistic evaluates 
the dependency of a covariate on the goodness of fit.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
statistic given all permutations of the response variable can be calculated and used to standardize 
the test statistic. In order to map multivariate test statistics into the real line two options are 
given: either the maximum of the absolute values of the standardized statistic or a quadratic form 
is considered. The resulting values can then be transformed to the scale of p~-~values where they 
can be compared in an unbiased way to select a split variable.
In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
observed values or optionally at an average value lying between two observations. To select the
best split point two-sample linear statistics of the same structure as in the independence test 
are evaluated for all possible subsets of the sample space of the selected split variable.

\subsubsection{MOB}\label{sec:mob}
While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
was developed with the aim of embedding parametric models into the framework of regression trees
and is hence purely intended for a model-based approach.

An objective function $\psi(Y,X,\theta)$ depending on observations $(Y,X)$, with a response $Y$ and 
covariate(s) $X$, and a $k$-dimensional vector of parameters $\theta$ is minimized to obtain the 
parameter estimate $\hat{\theta}$ for the considered parametric model fitted in the current node of 
the tree. For this purpose many different M-type estimators can be applied including maximum likelihood 
or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
algorithm can be employed as objective function. 
To obtain additional information on the goodness of fit of the parametric model so~-~called scores are
calculated. Scores are the values of the derivative of the objective function with respect to the parameters 
with the estimated parameters and the observed values plugged in:
$$
S(y, x, \hat{\theta}) = \frac{\partial \psi(y, x, \hat{\theta})}{\partial \theta}
$$
which is a $n \times k$ - matrix for a $k$-dimensional parameter vector $\theta$ and
a data set of $n$ observations containing the response vector $y = (y_1,\ldots,y_n)^\top$
and regressor(s) $x = (x_1,\ldots,x_n)^\top$. 
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$. 
After calculating these scores the aim is to find out whether there is any instability in the parameters
over each particular ordering induced by the possible partitioning variables. If a covariate has
no specific influence on the parameters the scores would always fluctuate randomly around their 
mean 0. Therefore, these deviations are investigated by measuring them using an empirical fluctuation 
process. Applying a scalar function which captures the fluctuation in this process yields the
test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
In this parametric unconditional framework p~-~values can be calculated and the covariate
with the smallest p~-~value is selected as split variable. Then an exhaustive search over all possible
partitions can be applied to find the best split point. Hence, the number of possible breakpoints is
not fixed unlike in the CTree algorithm which can make this operation computationally expensive.

\subsubsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
From the model fitted to the data in the current node residuals are calculated. But before applying
a statistical test two additional steps are performed (to prepare(?) the data):
The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
and for each possible split variable the observed values are categorized into 4 groups 
separated by the corresponding 1st, 2nd and 3rd quartile.
Then, a $\chi^2$ test of independence is performed for the binarized residuals and
each categorized split variable.
After choosing the split variable showing the highest dependency by yielding the lowest p~-~value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables such that the quartiles are
the only three possible split points.

@@

\textit{Note:}
While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
independence tests such that no split is performed if the lowest p~-~value is not smaller than 
the significance level.

@@

\subsubsection{Residuals vs full model scores}
Following the basic tree building strategy in Section~\ref{sec:urp}, a $n \times k$ - matrix is 
obtained based on the discrepancy measure. In the here discussed unifying framework this matrix 
is the score matrix $S$ as presented in Section~\ref{sec:mob} providing information on the 
goodness of fit for each observation and model parameter.
Since the first column of the score matrix is the vector of residuals the decision of using either 
residuals or full model scores for the test statistic is equivalent to the decision of using only 
the first column or the whole matrix $S$.

By including all available scores the goodness of fit of the model is investigated regarding each 
parameter rather than just regarding the mean/location parameter. In that way changes in parameters
other than the mean/location parameter should be detected more easily.

\textit{Example LM:} To exemplify this feature a linear model tree is considered as this allows 
for an easy and well structured description of the investigated subjects/issues. Therefore, from
here on this example will be used for illustrations.
(However, the conclusions drawn for this example also apply for a wide range of other model
types.)
The models fitted in the nodes of the considered linear tree model are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
The coefficient vector $\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$ is estimated
via minimizing the residual sum of squares (RSS) for the learning data set in 
the current node, for notational simplicity generally denoted by 
$\{(y_i, x_i)\}_{i=1\dots n}$. Hence, the discrepancy measure, i.e. the objective function is
$$
RSS(y_i, x_i, \boldsymbol{\beta}) = \sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
$$ 
such that the corresponding score matrix is defined by
$$
s(y_i, x_i, \hat{\boldsymbol{\beta}}) = \frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
$$
which is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is two dimensional. 
(In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
and $j \in \{1,2\}$.)
In this example considering only residuals corresponds to including the derivatives of $RSS$
only regarding the intercept parameter $\beta_0$, i.e. the first column of $S$.
On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
$\beta_1$ (second column).

\subsubsection{Binarization of scores}
Rather than applying the independence test directly on the selected (columns of) scores a
binarized version of them can be used as well. In particular, the score values are binarized
at 0 such that only their signs are considered.

\subsubsection{Categorization of covariates}
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd and 
3rd quartile are used to bin each possible split variable into four categories.

%\subsubsection{Conditional vs unconditional}

\begin{table}[t!]
\centering
\caption[Table caption text]{CTree, MOB, and GUIDE with the corresponding setting of
the switches in the unifying framework and the approximation type of the underlying null distribution.}
\label{tab:switches}
\begin{tabular}{ l c c c c }
\hline 
& Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & off           & off             & conditional\\
MOB   & full model scores & off           & off             & unconditional\\
GUIDE & residuals         & on            & on              & unconditional\\
\hline
\end{tabular}
\end{table}

@@

%All three algorithms apply statistical tests yielding p~-~values based on which
%the most appropriate split variable is chosen. One main difference lies in
%the way how scores are included. While CTree and MOB directly include the values
%of all available scores the GUIDE algorithm only uses residuals, in particular,
%the signs of residuals and binned versions of the covariates.
%The advantages of including all scores or only residuals as well as the impact
%of the binarization of scores at zero and the binning of covariates is
%investigated in detail in Section~\ref{sec:results}.
%Hence, the GUIDE algorithm differs mainly in the setting of the applied test
%in which the MOB and CTree algorithms are very similar. However, one essential difference
%between the testing strategies of MOB and CTree lies in the underlying approximation 
%of the null distribution where MOB employs an unconditional approach contrary
%to the conditional framework of CTree.

@@

par(mfrow = c(3,3))

>>>>

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
source("../guidelike.R")
source("../tests.R")
#load("../sim/simres20180422_1step.rda")
load("../sim/simres20181223_stump.rda")
sim_stump <- simres$res
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
load("../sim/simres20181223_tree.rda")
sim_tree <- simres$res
rm(simres)

## HCL palette
pal <- hcl(c(140, 80, 260, 70, 10), 300, 60)
names(pal) <- c("CTree", "CTree_max", "MOB", "GUIDE", "GUIDE_full")

@@

\Abstract{
A core step of every algorithm for learning regression trees is the decision if, 
how, and where to split the underlying data - or, in other words, the selection 
of the best splitting variable from the available covariates and the corresponding 
split point. Early tree algorithms (e.g., AID, CART) employ greedy search strategies 
that directly compare all possible split points in all available covariates. However, 
subsequent research showed that this is biased towards selection of covariates with 
more potential split points. Therefore, unbiased recursive partitioning algorithms 
have been suggested (e.g., QUEST, GUIDE, CTree, MOB) that first select the covariate 
based on statistical inference using $p$-values that are appropriately adjusted for 
the possible split points. In a second step a split point optimizing some objective 
function is selected in the chosen split variable. However, different unbiased tree 
algorithms employ different inference frameworks for computing these $p$-values and 
their relative advantages or disadvantages are not well understood, yet.

Therefore, three different approaches are considered here and embedded into a common 
modeling framework with special emphasis to linear model trees: classical categorical 
association tests (GUIDE), conditional inference (CTree), parameter instability 
tests (MOB). It is assessed how different building blocks affect the power of the 
tree algorithms to select the appropriate covariates for splitting: residuals vs. full 
model scores, binarization of residuals/scores at zero, binning of covariates, 
conditional vs.\ unconditional approximations of the null distribution.
}
\Keywords{classification and regression trees, independence tests, recursive partitioning}

@@

\section{Introduction}\label{sec:introduction}
In many situations fitting one global model to a data set can be very challenging, 
especially if this data contains lots of different features with high/extreme variations 
or complex interactions. Therefore, separating the data into more homogeneous subgroups 
based on (a set of) covariates first can simplify the task and fitting a local model to 
each of the resulting subgroups often leads to better results. This separation of the 
data can be done by applying a tree algorithm. Many different algorithms have been developed
in this research field. While they all follow the general idea of splitting the data such 
that some objective function is minimized/maximized, they differ in their 
specific/particular approaches to selecting a split variable and the corresponding split point. 
Some of the first tree algorithms (e.g., AID \citealp{Morgan+Sonquist:1963}; 
CART \citealp{Breiman+Friedman+Stone:1984}) rely on exhaustive search procedures to find both, 
the best split point and split variable in one step by directly comparing all possible split
points in all possible split variables. However, it has been shown that this is not only
computationally expensive but also biased towards split variables with many possible split
points \citep{Doyle:1973, Kim+Loh:2001}. 
Therefore, selecting a split variable in a first step and then searching for the 
best split point only within this variable in a separated second step is a more 
promising strategy as applied for example by the algorithms QUEST \citep{Loh+Shih:1997},
CHAID \citep{Kass:1980}, GUIDE \citep{Loh:2002}, CTree \citep{Hothorn+Hornik+Zeileis:2006} 
and MOB \citep{Zeileis+Hothorn+Hornik:2008}.
For the first step of selecting a split variable they all follow the idea of 
choosing the covariate which shows the highest association to/influence on the response variable 
based on a statistical test. While QUEST, CHAID and GUIDE employ statistical significance
tests for contingency tables, CTree applies permutation tests in a conditional inference
framework and MOB uses fluctuation tests based on asymptotic theory.

In this paper the focus will be put on this first step, i.e. the task of selecting 
the right/best split variable. In particular, the approach of the GUIDE algorithm is
compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
unifying framework for unbiased model based tree algorithms is presented such that each 
of the three algorithms can be obtained by a specific combination of the available building 
blocks. In that way, the effects of each of these characteristic features of the GUIDE,
CTree and MOB algorithm can be investigated separately by varying the switches in
the tool box provided by this general framework. 
Depending on the setup different effects can be observed for the investigated characteristics.
However, none of the results contradicts the conclusion that in general including all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated and the impact of applying a conditional
or an unconditional approximation of the null distribution is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated testing strategies more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for model 
based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

\section{Unbiased Recursive Partitioning}
\label{sec:urp}

\subsection{Generic Algorithm}
The basic idea of building a regression tree model is to partition the data into smaller and 
more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
have been developed, all following the same general structure:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
%This can for example be the sum of deviations from a typical or average value of the 
%data set or it can be based on a fitted model.
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data the discrepancy measure is minimal within the resulting subgroups.
\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criterion
is reached or the subgroup consists of only one or all equal observations. (In the second
case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
\end{enumerate}

@@

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical (independence) tests
following the same general strategy: 
Based on the discrepancy measure an $n \times k$ - matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
necessary to calculate the corresponding $p$-values in order to allow for an unbiased
variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
$p$-value is selected as split variable as it has shown to be the variable with the strongest
influence on the goodness of fit of the model.
% Hinweis: pruning-Strategien
%Once the split variable is chosen the corresponding split point is then selected such that it 
%minimizes the discrepancy measure.

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Sections~\ref{sec:ctree}.

@@

\subsection{CTree}\label{sec:ctree}

The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
a conditional inference framework by applying permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split variable $Z$.
For different types of response variables different functions $h$ which depend on $Y$ in a 
permutation-symmetric way and might also be multidimensional can be selected.
For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
response can be mapped to a unity vector indicating the corresponding category by an indicator
function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$. Alternatively, the function $h$ can return a location
and scale parameter of $Y$, e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
the corresponding model scores can be obtained by setting $h(Y) = s(Y,X,\hat{\theta})$
where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function:
$$
s(Y, X, \hat{\theta}) = \frac{\partial \psi(Y, X, \hat{\theta})}{\partial \theta}
$$
with $\psi$ being the objective function which is minimized to estimate the model parameter
$\theta$. Hence, for a data set of $n$ observations containing the response vector 
$y = (y_1,\ldots,y_n)^\top$ and regressor(s) $x = (x_1,\ldots,x_n)^\top$ and a $k$-dimensional 
parameter vector $\theta$ the score function $s$ returns a $n \times k$ - matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$. 
(To estimate $\theta$ many different M-type estimators can be applied including maximum likelihood 
or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
algorithm can be employed as objective function.)

Similarly, different types of functions can be chosen for the influence function $g$ depending
on a possible split variable $Z$. A simple choice in case of $Z$ being a numeric variable is again
the identity function $g(Z) = Z$. For categorical variables $g$ can also map its values to the 
corresponding unity vectors by $g(Z) = (0,\ldots,1,\ldots,0)$.

To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the test 
statistic given all permutations of the response variable can be calculated and used to standardize 
the test statistic. In order to map multivariate test statistics into the real line two options are 
given: either the maximum of the absolute values of the standardized statistic or a quadratic form 
is considered. The resulting values can then be transformed to the scale of $p$-values where they 
can be compared in an unbiased way to select a split variable.

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree algorithm
provides a very general framework where the choice of the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test.

In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
observed values or optionally at an average value lying between two observations. To select the
best split point two-sample linear statistics of the same structure as in the independence test 
are evaluated for all possible subsets of the sample space of the selected split variable.

\subsection{MOB}\label{sec:mob}
While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
was developed with the aim of embedding parametric models into the framework of regression trees
and is hence purely intended for a model-based approach. The testing strategy for selecting a split
variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
node of the tree, and on the application of the Central Limit Theorem. However, the fitted parametric 
models can be of a very basic and simple form such as an average value. The algorithm only requires the 
corresponding model scores $s(Y,X,\hat{\theta})$ in order to test for any instability in the model 
parameters over each particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the scores would always fluctuate randomly
around their mean 0. Therefore, these deviations are investigated by measuring them using an 
empirical fluctuation process. Applying a scalar function which captures the fluctuation in this process 
yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
the resulting test statistic asymptotic theory, in particular the Central Limit Theorem is applied.
In this parametric unconditional framework $p$-values can be calculated and the covariate
with the smallest $p$-value is selected as split variable. 
For $Z$ being a numeric variable a maximally selected split point search is employed while for a 
categorical variable all possible splits are evaluated to find the best one.
Hence, the number of possible breakpoints is not fixed unlike in the CTree algorithm which can make 
this operation computationally expensive.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
For the statistical test applied to select a split variable in each node it requires some type of 
residuals $r(Y,X,\hat{\theta})$ from the models such as for example 
$r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
Before applying a statistical test two additional steps are performed:
The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
and for each possible split variable the observed values are categorized into 4 groups 
separated by the corresponding 1st, 2nd and 3rd quartile.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized split variable.
After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables such that the quartiles are
the only three possible split points.

@@

\textit{Note:}
While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
independence tests such that no split is performed if the lowest $p$-value is not smaller than 
the significance level. The impact of the choice of pruning strategy is further discussed in 
Section~\ref{sec:discussion}.

@@

\subsection{Switches}
\begin{itemize}
\item  Transformations of Y:\\
The selection of a split variable is based on a statistical test investigating the dependency of the
response $Y$ on each possible split variables $Z$. However, before evaluating such a test a model-based
transformation of $Y$ is applied to obtain a discrepancy measure. This can be residuals 
$r(Y,X,\hat{\theta})$ as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm
or, more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
hence the full model matrix $S$.
%providing information on the goodness of fit for each observation and model parameter.
(By including all available scores the goodness of fit of the model is investigated regarding 
each parameter rather than just regarding the mean/location parameter. In that way changes in 
parameters other than the mean/location parameter should be detected more easily.)

\item Binarization of scores:\\
Rather than applying the independence test directly on the selected (columns of) scores a
binarized version of them can be used as well. In particular, the score values are binarized
at $0$ such that only their signs are considered.

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are used to bin each possible split variable into four categories. In that way
a $\chi^2$-test can be applied and the quartiles are the only possible split points while otherwise maximally selected splits over $Z$ are considered.

\end{itemize}
%\subsubsection{Conditional vs unconditional}

\begin{table}[t!]
\centering
\caption[Table caption text]{CTree, MOB, and GUIDE with the corresponding setting of
the switches in the unifying framework and the approximation type of the underlying null distribution.}
\label{tab:switches}
\begin{tabular}{ l c c c c }
\hline 
& Scores            & Binarization  & Categorization  & Null distribution \\ 
\hline
CTree & full model scores & -           & -             & conditional\\
MOB   & full model scores & -           & -             & unconditional\\
GUIDE & residuals         & \checkmark  & \checkmark    & unconditional\\
\hline
\end{tabular}
\end{table}

\subsection{Linear model tree}
As an example a linear model tree is considered as this allows 
for an easy and well structured description of the investigated subjects/issues. 
For that reason, this model is also considered for the simulation study in Section~\ref{sec:simulation}.
However, the conclusions drawn for this example also apply for a wide range of other model
types. The models fitted in the nodes of the considered linear tree model are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
As for each node the type of the fitted model is the same no node-specific features are considered
and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepency measure can be obtained by simply calculating deviations of the 
response values from the mean value within the node, for example by $\sum_{i=1}^n (\bar{y} - y_i)^2$.
However, this includes only the response variable $Y$ but not the fitted model.
For a model-based approach residual can be considered such as in the residual sum of squares (RSS)
$$
r(y,x,\boldsymbol{\beta}) = RSS(y, x, \boldsymbol{\beta}) = 
\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2.
$$ 
which is also the objective function that is minimized for the estimation of the coefficient vector 
$\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding scores are defined by
$$
s(y_i, x_i, \hat{\boldsymbol{\beta}}) = 
\frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
$$
such that the score matrix $S$ is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is 
two dimensional. In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
and $j \in \{1,2\}$.

In this example considering only residuals, i.e. only the first column of the score matrix $S$,
corresponds to including the derivatives of $RSS$ only regarding the intercept parameter~$\beta_0$.
On the contrary, using full model scores means including the whole $n \times 2$ matrix $S$, hence,
the derivatives of $RSS$ regarding the intercept $\beta_0$ (first column) and the slope parameter 
$\beta_1$ (second column).

@@

%All three algorithms apply statistical tests yielding $p$-values based on which
%the most appropriate split variable is chosen. One main difference lies in
%the way how scores are included. While CTree and MOB directly include the values
%of all available scores the GUIDE algorithm only uses residuals, in particular,
%the signs of residuals and binned versions of the covariates.
%The advantages of including all scores or only residuals as well as the impact
%of the binarization of scores at zero and the binning of covariates is
%investigated in detail in Section~\ref{sec:results}.
%Hence, the GUIDE algorithm differs mainly in the setting of the applied test
%in which the MOB and CTree algorithms are very similar. However, one essential difference
%between the testing strategies of MOB and CTree lies in the underlying approximation 
%of the null distribution where MOB employs an unconditional approach contrary
%to the conditional framework of CTree.

@@
